---
title: "Week 13"
subtitle: "Communicating Results II"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Announcements

### Upcoming schedule

What's left:

+ Lab 09 due Thursday (last one!) - "presentation ready" means it incorporates the tips we talked about last week! 
    + Declutter - reduce to essentials
    + What's your story? What data tells it?
    + Choose a visual to make your key point / comparison salient
    + Use highlights, annotations, and informative titles to SHOW your story
+ Quiz 07 Thursday - communicating results

### Project deadlines / important dates

+ Th April 16 (11:59pm): Presentation rough draft
+ Tue April 21 (4pm): Written report rough draft - BRING 3 HARD COPIES TO CLASS WITH YOU! Must be on time, must read assigned groups' paper before class Thursday
+ Tue April 21 (in-class): Peer review of presentations - must be in class!
+ Th April 23 (in class): Peer review of rough drafts - must be in class!
+ Sat April 25 (11:59pm): Formal peer review submitted
+ Tue & Th April 28 & 30: Presentations - must be in class both days!
+ Monday May 4: Final deliverables (report + repository)

Upcoming deadlines are hard deadlines - your timeliness affects your peers! Points are associated with all of the above.

```{r}
groups <- c("Hannah & Briell", "Sophie & Claire", "Catherine, Deana & Perry",
            "Jay & Ryan", "Alex & Andrew", "Will & Zach", "Mei & Nixie")

set.seed(437)
sample(groups)
```

+ Presentations will be 10-12 minutes each, with 5 minutes for questions. 
+ Will be required to submit at least 1 question per presentation (have to be present for both class sessions)

## Today: Project work session

### Part 1: brainstorm in your project groups (10 mins)

+ What is our story missing at the moment? 
+ What additional analyses need to be conducted?
+ What are possible counter-arguments to our story?

\
\
\
\
\
\
\
\

### Part 2: have AI stress-test your argument (10 minutes)

+ Ask AI for feedback / stress test your argument! 
    + What new questions could be answered with my existing data that I haven’t considered? Why do they matter?
    + Which variables or patterns in my data deserve deeper analysis? What relationships should I explore further?
    + What am I missing? Where is my evidence insufficient, misinterpreted, or potentially biased?
    + What are the weakest parts of my argument and why?
    + What are the strongest counter-arguments to my claims? Rank them by how difficult they are to respond to. 
    + How would someone with a different worldview or who fundamentally disagrees with my assumptions interpret this argument? What would they say I'm overlooking?
+ What will you do / add / change in light of this brainstorming & stress testing? 
    
### Part 3: planning & work session
 
+ Create a to-do list - be specific, and include a timeline! 
    + Printed version of project website info provided
+ Start by creating a to-do list - be specific, and include a timeline! 
    + Printed version of project website info provided
+ What updates do you need to make to your visualizations in light of Thursday's lecture? What new visualizations / iterations should you try? Don't get pigeon holed into your first idea.
    + [`ggtext` package](https://r-graph-gallery.com/package/ggtext.html) to add color to your plot titles
    + Make your font sizes much bigger than you think you need to! Your audience will be viewing from the back of the room. 
+ Identify 3 - 5 (or more) external sources that you will incorporate into your data story. To provide background, context, supplementary evidence, etc.
    + You are not the first person to investigate this issue! 
    + What have others had to say about it? 

<!-- ### Citation management -->

<!-- -   Download & Install [LaTeX](https://www.latex-project.org/get/#tex-distributions) -->

<!-- -   Download & Install [Zotero](https://www.zotero.org) -->

<!-- -   Install the [Zotero Connector for Chrome](https://www.zotero.org/download/connectors) -->

<!-- -   Install [Better BibTeX for Zotero](https://retorque.re/zotero-better-bibtex/installation/) -->

<!-- -   See `Zotero_demo.qmd` on GitHub (and `references.bib`) for demo of citations in Quarto -->

<!-- -   Open Microsoft Word on your computer and look for the [Zotero tab](https://www.zotero.org/support/word_processor_plugin_usage) -->

<!-- -   Open a Google Doc (in Chrome) and look for the [Zotero tab](https://www.zotero.org/support/google_docs) -->

<!-- Citation tasks: -->

<!-- 1.  In Chrome, search for and locate the sources you plan to cite, and click on the Extensions puzzle piece to add them to your Zotero library -->

<!--     1.  If they are just websites rather than academic papers or books, you may need to manually edit the citation info in Zotero -->

<!-- 2.  In a Quarto file (e.g. your written report), create (potentially placeholder) sentences with in-text citations (Insert \> \@ Citation), then Render. -->

<!-- 3.  OPTIONAL: Repeat the citation process in Word (or Google Docs) -->

<!-- ### Thursday: Rough Draft Peer Review -->

<!-- Why peer review? -->

<!-- -   Giving good feedback is a skill (that we don't often practice) -->

<!-- -   Receiving and incorporating feedback effectively also takes practice -->

<!-- -   Our work is better when we have multiple perspectives speak into it -->

<!-- -   It's one of the only cures to the curse of expertise! -->

<!-- -   Makes us better at self-assessing our own work -->

<!-- -   It's part of the scientific process -->

<!-- ## Logistics:  -->

<!-- -   In your project groups, discuss assigned paper and fill out the "FORMAL Rough Draft Peer Feedback - Group X" form in the shared Google Drive folder (located in the folder of the group you're reviewing) -->

<!-- -   Then, you will write up a 1-2 page formal review, which will include: -->

<!--     -   A summary of the research questions, data context, and findings, as you understand them -->

<!--     -   Brief comments on strengths of the paper -->

<!--     -   Constructive criticism, raising questions / points of confusion, and outlining what needs to be changed/improved in order for the report to be exemplary -->

<!-- -   The constructive criticism should make up the bulk of the 1-page review. Remember: the goal is to help them IMPROVE the work. You should NOT simply tell them it's wonderful, and you should NOT simply criticize without offering concrete suggestions for HOW to improve. -->

<!-- ###  -->
