---
title: "Week 12"
subtitle: "Project / Exam 2"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Statistician(s) of the Week

[Jo Hardin](https://hardin47.netlify.app/about/)

[Mine Cetinkaya-Rundel](https://mine-cr.com)

### Upcoming schedule

What's left:

+ You've finished annotations - thanks for excellent engagement :) 
+ Lab 09 due next Thursday (started today)
+ Quiz 07 next Thursday - communicating results
+ Presentation rough draft (Th April 16)
+ Written report rough draft (Tue April 21)
+ Formal peer review (Sat April 25)
+ Final deliverables (report + repository) (Mon May 4)

The last four bullet points are hard deadlines - your timeliness affects your peers! 

| Week | Tue in-class | Thur in-class |
|-----------|----------------|---------------|
| Week 12 (April 7, 9) | Web scraping | Communicating Results |
| Week 13 (April 14, 16) | Communicating results / Project work session | Quiz 07 + Project work session |
| Week 14 (April 21, 23) | Peer review (presentations) | Peer review (written reports) |
| Week 15 (April 28, 30) | Presentations | Presentations |

```{r}
groups <- c("Hannah & Briell", "Sophie & Claire", "Catherine, Deana & Perry",
            "Jay & Ryan", "Alex & Andrew", "Will & Zach", "Mei & Nixie")

set.seed(437)
sample(groups)
```

+ Presentations will be 10-12 minutes each, with 5 minutes for questions. 
+ Will be required to submit at least 1 question per presentation (have to be present for both class sessions)

### Questions?

<!-- ### Rough Draft Peer Review -->

<!-- Why peer review? -->

<!-- -   Giving good feedback is a skill (that we don't often practice) -->

<!-- -   Receiving and incorporating feedback effectively also takes practice -->

<!-- -   Our work is better when we have multiple perspectives speak into it -->

<!-- -   Makes us better at self-assessing our own work -->

<!-- -   It's part of the scientific process -->

<!-- Logistics: -->

<!-- -   (15 mins) In your project groups, discuss assigned paper and fill out the "SHORT Rough Draft Peer Feedback - Group X" form in the shared [Google Drive folder](https://drive.google.com/drive/folders/190TgJca5qleOGoQQMJ6IbLj_RbSRKtOE?usp=share_link) (located in the folder of the group you're reviewing) -->

<!-- -   (15 mins) Pair up with the group and exchange feedback (\~7 mins per group) -->

<!-- -   Remainder of class will be spent on working on your FORMAL review (this is graded and counts towards your final project grade!) -->

<!-- +---------------+--------------------+ -->
<!-- |               | Group pairings     | -->
<!-- +===============+====================+ -->
<!-- | Session 1     | Group 1 & Group 3  | -->
<!-- |               |                    | -->
<!-- |               | Group 2 & Group 4  | -->
<!-- |               |                    | -->
<!-- |               | Group 5 & Dr. Fitz | -->
<!-- +---------------+--------------------+ -->

<!-- For the group you did NOT already exchange feedback with, you will write up a FORMAL peer review -->

<!-- -   Group 1 formally reviews Group 2 -->
<!-- -   Group 2 formally reviews Group 3 -->
<!-- -   Group 3 formally reviews Group 4 -->
<!-- -   Group 4 formally reviews Group 5 -->
<!-- -   Group 5 formally reviews Group 1 -->

<!-- Your formal review will include: -->

<!-- -   FORMAL evaluation rubric filled out (Written report section ONLY) (in the folder of the group you're reviewing) -->
<!-- -   A brief write-up (\~1 page) highlighting the following: -->
<!--     -   A summary of the research questions, data context, and findings, as you understand them -->
<!--     -   Brief comments on strengths of the paper -->
<!--     -   Constructive criticism, raising questions / points of confusion, and outlining what needs to be changed/improved in order for the report to be exemplary -->

<!-- The constructive criticism should make up the bulk of the 1-page review. Remember: the goal is to help them IMPROVE the work. You should NOT simply tell them it's wonderful, and you should NOT simply criticize without offering concrete suggestions for HOW to improve. -->
