---
title: "Data Wrangling Ib"
subtitle: "STAT 4380"
author: "Katie Fitzgerald, adpated from datasciencebox.org"
format:
  revealjs:
    theme: stat4380-slides.css
    width: 1280
    height: 720
    margin: 0
    slide-number: false
    progress: false
    toc: false
    center: false
    transition: none
    background-transition: none
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE,
  dplyr.print_min = 6,
  dplyr.print_max = 6,
  tibble.width = 65,
  width = 65
)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.asp = 0.618,
  out.width = "60%",
  fig.align = "center",
  dpi = 300,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
```

## More `dplyr` verbs for wrangling a single data frame {.section-divider .middle}

---

## We... {.section-divider .middle}

::: {.hand}
[have]{.huge .green} a single data frame

[want]{.huge .pink} to slice it, dice it, juice it, and process it
:::

---

## Data: Hotel bookings

- Data from two hotels: one resort and one city hotel
- Observations: Each row represents a hotel booking

```{r message=FALSE}
hotels <- read_csv("data/hotels.csv")
```

---

## `select`, `arrange`, and `slice` {.section-divider .middle}

---



## `filter` {.section-divider .middle}

---

## `filter` to select a subset of rows

::: {.midi}
```{r output.lines=17}
#| code-line-numbers: "3"
# bookings in City Hotels
hotels |>
  filter(hotel == "City Hotel")
```
:::

---

## Quiz {.section-divider .middle}

What rows remain? 

```{r, eval = FALSE}
df <- data.frame(student = c("A","B","C","D"),
             score = c(90, 80, 85, 95))
df |>  filter(score > 85)
```

    A. Students A and D
    B. Students B and C
    C. Students A, C, and D
    D. All students

---

## `filter` for many conditions at once

```{r}
#| code-line-numbers: "3-4"
hotels |>
  filter( 
    adults == 0,
    children >= 1
    ) |> 
  select(adults, babies, children)
```

---

## `filter` for more complex conditions

```{r}
#| code-line-numbers: "5"
# bookings with no adults and some children or babies in the room
hotels |>
  filter( 
    adults == 0,     
    children >= 1 | babies >= 1     # | means or
    ) |>
  select(adults, babies, children)
```

---

## Logical operators in R

<br>

## Logical operators in R

<br>

::: {.columns}

::: {.column width="50%"}
| operator | definition |
|---|---|
| `<` | less than |
| `<=` | less than or equal to |
| `>` | greater than |
| `>=` | greater than or equal to |
| `==` | exactly equal to |
| `!=` | not equal to |

:::

::: {.column width="50%"}
| operator | definition |
|---|---|
| `x & y` | `x` AND `y` |
| `x | y` | `x` OR `y` |
| `is.na(x)` | test if `x` is `NA` |
| `!is.na(x)` | test if `x` is not `NA` |
| `x %in% y` | test if `x` is in `y` |
| `!(x %in% y)` | test if `x` is not in `y` |
|  |  |
:::

:::
---

## `distinct` and `count` {.section-divider .middle}

---

```{r include=FALSE}
options(dplyr.print_max = 20)
```


## `distinct` to filter for unique rows

... and `arrange` to order alphabetically

::: {.small}
::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "2"
hotels |> 
  distinct(market_segment) |>
  arrange(market_segment)
```
:::
::: {.column width="50%"}
```{r output.lines=13}
#| code-line-numbers: "2"
hotels |> 
  distinct(hotel, market_segment) |>
  arrange(hotel, market_segment)
```
:::
:::
:::

---

## `count` to create frequency tables

::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "3"
# alphabetical order by default
hotels |>
  count(market_segment)
```
:::
::: {.column width="50%" .fragment}
```{r}
#| code-line-numbers: "3"
# descending frequency order
hotels |>
  count(market_segment, sort = TRUE)
```
:::
:::

---

## `count` and `arrange`

::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "4"
# ascending frequency order
hotels |>
  count(market_segment) |>
  arrange(n)
```
:::
::: {.column width="50%"}
```{r}
#| code-line-numbers: "5"
# descending frequency order
# just like adding sort = TRUE
hotels |>
  count(market_segment) |>
  arrange(desc(n))
```
:::
:::

---

## `count` for multiple variables

```{r}
#| code-line-numbers: "2"
hotels |>
  count(hotel, market_segment)
```

---

## order matters when you `count`

::: {.midi}
::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: "3"
# hotel type first
hotels |>
  count(hotel, market_segment)
```
:::
::: {.column width="50%"}
```{r}
#| code-line-numbers: "3"
# market segment first
hotels |>
  count(market_segment, hotel)
```
:::
:::
:::

---


## `mutate` {.section-divider .middle}

---

## `mutate` to add a new variable

```{r}
#| code-line-numbers: "2"
hotels |>
  mutate(little_ones = children + babies) |>
  select(children, babies, little_ones) |>
  arrange(desc(little_ones))
```

---

## Little ones in resort and city hotels

::: {.midi}
::: {.columns}
::: {.column width="50%"}
```{r}
# Resort Hotel
hotels |>
  mutate(little_ones = children + babies) |>
  filter(
    little_ones >= 1,
    hotel == "Resort Hotel"
    ) |>
  select(hotel, little_ones)
```
:::
::: {.column width="50%"}
```{r}
# City Hotel
hotels |>
  mutate(little_ones = children + babies) |>
  filter(
    little_ones >= 1,
    hotel == "City Hotel"
    )  |>
  select(hotel, little_ones)
```
:::
:::
:::

---

::: {.question}
What is happening in the following chunk?
:::

::: {.midi}
```{r}
hotels |>
  mutate(little_ones = children + babies) |>
  count(hotel, little_ones) |>
  mutate(prop = n / sum(n))
```
:::

---

## `summarise` and `group_by` {.section-divider .middle}

---

## `summarise` for summary stats

```{r}
#| code-line-numbers: "3"
# mean average daily rate for all bookings
hotels |>
  summarise(mean_adr = mean(adr))
```

::: {.fragment}
::: {.tip}
`summarise()` changes the data frame entirely, it collapses rows down to a single 
summary statistic, and removes all columns that are irrelevant to the calculation.
:::
:::

---

::: {.tip}
`summarise()` also lets you get away with being sloppy and not naming your new 
column, but that's not recommended!
:::

::: {.columns}
::: {.column width="50%"}
`r emo::ji("x")`

```{r}
hotels |>
  summarise(mean(adr))
```
:::
::: {.column width="50%"}
`r emo::ji("white_check_mark")`

```{r}
hotels |>
  summarise(mean_adr = mean(adr))
```
:::
:::

---

## `group_by` for grouped operations

```{r}
#| code-line-numbers: "3"
# mean average daily rate for all booking at city and resort hotels
hotels |>
  group_by(hotel) |>
  summarise(mean_adr = mean(adr))
```

---

## Calculating frequencies

The following two give the same result, so `count` is simply short for `group_by` then determine frequencies 

::: {.columns}
::: {.column width="50%"}
```{r}
hotels |>
  group_by(hotel) |>
  summarise(n = n())
```
:::
::: {.column width="50%"}
```{r}
hotels |>
  count(hotel)
```
:::
:::

---

## Multiple summary statistics

`summarise` can be used for multiple summary statistics as well

```{r}
hotels |>
  summarise(
    min_adr = min(adr),
    mean_adr = mean(adr),
    median_adr = median(adr),
    max_adr = max(adr)
    )
```

---

## Quiz {.section-divider .middle}

What does this code return?

```{r, eval = FALSE}
df |> 
  summarise(mean_score = mean(score))
```

A. A 1x1 data frame with one value: the mean of all scores

B. A new column with each student’s mean score

C. An error: `summarise()` needs `group_by()`

D. The same dataset as before

---

## Quiz {.section-divider .middle}

What's the result of this pipeline?

```{r, eval = FALSE}
df %>%
  mutate(passed = score > 85) %>%
  filter(passed)
```


A. Keeps only students with score > 85

B. Creates a new column passed with TRUE/FALSE for all students

C. Both A and B

D. Neither
