---
title: "Data Wrangling Ib"
subtitle: "STAT 4380"
author: "Katie Fitzgerald, adpated from datasciencebox.org"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: middle

# More `dplyr` verbs for wrangling a single data frame

---

class: middle

# .hand[We...]

.huge[.green[have]] .hand[a single data frame]

.huge[.pink[want]] .hand[to slice it, dice it, juice it, and process it]

---

## Data: Hotel bookings

- Data from two hotels: one resort and one city hotel
- Observations: Each row represents a hotel booking

```{r message=FALSE}
hotels <- read_csv("data/hotels.csv")
```

---

class: middle

# `select`, `arrange`, and `slice`

---



class: middle

# `filter`

---

## `filter` to select a subset of rows

.midi[
```{r output.lines=17}
# bookings in City Hotels
hotels |>
  filter(hotel == "City Hotel") #<<
```
]

---

# Quiz

What rows remain? 

```{r, eval = FALSE}
df <- data.frame(student = c("A","B","C","D"),
             score = c(90, 80, 85, 95))
df |>  filter(score > 85)
```

    A. Students A and D
    B. Students B and C
    C. Students A, C, and D
    D. All students

---

## `filter` for many conditions at once

```{r}
hotels |>
  filter( 
    adults == 0,     #<<
    children >= 1    #<<
    ) |> 
  select(adults, babies, children)
```

---

## `filter` for more complex conditions

```{r}
# bookings with no adults and some children or babies in the room
hotels |>
  filter( 
    adults == 0,     
    children >= 1 | babies >= 1     # | means or  #<<
    ) |>
  select(adults, babies, children)
```

---

## Logical operators in R

<br>

operator    | definition                   || operator     | definition
------------|------------------------------||--------------|----------------
`<`         | less than                    ||`x`&nbsp;&#124;&nbsp;`y`     | `x` OR `y` 
`<=`        |	less than or equal to        ||`is.na(x)`    | test if `x` is `NA`
`>`         | greater than                 ||`!is.na(x)`   | test if `x` is not `NA`
`>=`        |	greater than or equal to     ||`x %in% y`    | test if `x` is in `y`
`==`        |	exactly equal to             ||`!(x %in% y)` | test if `x` is not in `y`
`!=`        |	not equal to                 ||`!x`          | not `x`
`x & y`     | `x` AND `y`                  ||              |

---

class: middle

# `distinct` and `count`

---

```{r include=FALSE}
options(dplyr.print_max = 20)
```


## `distinct` to filter for unique rows

... and `arrange` to order alphabetically

.small[
.pull-left[
```{r}
hotels |> 
  distinct(market_segment) |> #<<
  arrange(market_segment)
```
]
.pull-right[
```{r output.lines=13}
hotels |> 
  distinct(hotel, market_segment) |> #<<
  arrange(hotel, market_segment)
```
]
]

---

## `count` to create frequency tables

.pull-left[
```{r}
# alphabetical order by default
hotels |>
  count(market_segment) #<<
```
]
--
.pull-right[
```{r}
# descending frequency order
hotels |>
  count(market_segment, sort = TRUE) #<<
```
]

---

## `count` and `arrange`

.pull-left[
```{r}
# ascending frequency order
hotels |>
  count(market_segment) |>
  arrange(n) #<<
```
]
.pull-right[
```{r}
# descending frequency order
# just like adding sort = TRUE
hotels |>
  count(market_segment) |>
  arrange(desc(n)) #<<
```
]

---

## `count` for multiple variables

```{r}
hotels |>
  count(hotel, market_segment) #<<
```

---

## order matters when you `count`

.midi[
.pull-left[
```{r}
# hotel type first
hotels |>
  count(hotel, market_segment) #<<
```
]
.pull-right[
```{r}
# market segment first
hotels |>
  count(market_segment, hotel) #<<
```
]
]

---


class: middle

# `mutate`

---

## `mutate` to add a new variable

```{r}
hotels |>
  mutate(little_ones = children + babies) |> #<<
  select(children, babies, little_ones) |>
  arrange(desc(little_ones))
```

---

## Little ones in resort and city hotels

.midi[
.pull-left[
```{r}
# Resort Hotel
hotels |>
  mutate(little_ones = children + babies) |>
  filter(
    little_ones >= 1,
    hotel == "Resort Hotel"
    ) |>
  select(hotel, little_ones)
```
]
.pull-right[
```{r}
# City Hotel
hotels |>
  mutate(little_ones = children + babies) |>
  filter(
    little_ones >= 1,
    hotel == "City Hotel"
    )  |>
  select(hotel, little_ones)
```
]
]

---

.question[
What is happening in the following chunk?
]

.midi[
```{r}
hotels |>
  mutate(little_ones = children + babies) |>
  count(hotel, little_ones) |>
  mutate(prop = n / sum(n))
```
]

---

class: middle

# `summarise` and `group_by`

---

## `summarise` for summary stats

```{r}
# mean average daily rate for all bookings
hotels |>
  summarise(mean_adr = mean(adr)) #<<
```

--

.pull-left-wide[
.tip[
`summarise()` changes the data frame entirely, it collapses rows down to a single 
summary statistic, and removes all columns that are irrelevant to the calculation.
]
]

---

.tip[
`summarise()` also lets you get away with being sloppy and not naming your new 
column, but that's not recommended!
]

.pull-left[
`r emo::ji("x")`

```{r}
hotels |>
  summarise(mean(adr))
```
]
.pull-right[
`r emo::ji("white_check_mark")`

```{r}
hotels |>
  summarise(mean_adr = mean(adr))
```
]

---

## `group_by` for grouped operations

```{r}
# mean average daily rate for all booking at city and resort hotels
hotels |>
  group_by(hotel) |> #<<
  summarise(mean_adr = mean(adr))
```

---

## Calculating frequencies

The following two give the same result, so `count` is simply short for `group_by` then determine frequencies 

.pull-left[
```{r}
hotels |>
  group_by(hotel) |>
  summarise(n = n())
```
]
.pull-right[
```{r}
hotels |>
  count(hotel)
```
]

---

## Multiple summary statistics

`summarise` can be used for multiple summary statistics as well

```{r}
hotels |>
  summarise(
    min_adr = min(adr),
    mean_adr = mean(adr),
    median_adr = median(adr),
    max_adr = max(adr)
    )
```

---

# Quiz

What does this code return?

```{r, eval = FALSE}
df |> 
  summarise(mean_score = mean(score))
```

    A. A 1x1 data frame with one value: the mean of all scores
    B. A new column with each student’s mean score
    C. An error: `summarise()` needs `group_by()`
    D. The same dataset as before

---

# Quiz 

What's the result of this pipeline?

```{r, eval = FALSE}
df %>%
  mutate(passed = score > 85) %>%
  filter(passed)
```

    A. Keeps only students with score > 85
    B. Creates a new column passed with TRUE/FALSE for all students
    C. Both A and B
    D. Neither