---
title: "Inference for Numerical Data, Part I: One Mean"
subtitle: "MATH 250"
author: "Katie Fitzgerald, adpated from [Intro to Modern Statistics](https://openintro-ims.netlify.app)"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
library(dsbox)
library(scales)
library(infer)
library(distributional)
library(ggdist)
library(gridExtra)
library(openintro)
library(broom)
library(readxl)
```


class: middle

# Choosing the right analysis

---

## What do you want to do?

First step: Ask the following questions
  - What is the research question?
  - How many variables?
  - What type(s) of variable(s)?

--

Estimation -> Confidence interval

Decision -> Hypothesis test

--

Assumptions met?

---

class: middle

# Case study: Cherry Blossom Race

---

## Is the typical US runner getting faster or slower over time?

Cherry Blossom Race is a 10-mile race in Washington D.C each spring. In 2006, the average time for all runners was 93.29 minutes. 

We have data on 100 participants in the 2017 Cherry Blossom Race. 

```{r, echo = FALSE}
library(cherryblossom)
set.seed(437)
race17 <- run17 |> 
  sample_n(100)
```

--

+ How many variables? **1**
+ What type(s) of variable(s)? **numerical**

--

This is an **(dichotomous) decision** question, meaning a hypothesis test is most appropriate. 

--

The parameter of interest is the **population mean**, $\mu$, and the sample statistic is the sample mean, $\bar{x}$

---

## Hypothesis test for one mean

$$H_0: \mu_{2017} = 93.29$$
$$H_A: \mu_{2017} \neq 93.29$$
--

CLT tells us, $$T = \frac{\overline{x} - \mu}{SE(\overline{x})} = \frac{\overline{x} - \mu}{s/\sqrt{n}} \sim t_{n-1}$$

--

```{r, echo = FALSE}
race17 <- race17 |> 
  mutate(time_mins = net_sec/60)
```


```{r}
t_stat <- (mean(race17$time_mins) - 93.29)/(sd(race17$time_mins)/sqrt(nrow(race17)))
t_stat
```

**Assumption:** n > 30, independent observations
---

## Hypothesis test for one mean

.panelset[
.panel[.panel-name[Plot]
```{r, echo = FALSE, warning = FALSE, out.height = "60%"}
crit_value99 <- qt(.995, df = nrow(race17) - 1)
stat <-  abs((mean(race17$time_mins) -
                  93.29)/(sd(race17$time_mins)/sqrt(nrow(race17))))
ggplot() +
  xlim(c(-5,5)) + 
  stat_slab(aes(y = "T",
                   dist = dist_student_t(df = nrow(race17) - 1),
                   fill = after_stat(x < -stat | x > stat)),
               show.legend = FALSE) +
  geom_vline(xintercept = crit_value99, linewidth = 1, linetype = 21,
             color = "navyblue") +
  geom_vline(xintercept = -crit_value99, linewidth = 1, linetype = 21,
             color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = crit_value99, y = 2, color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = -crit_value99, y = 2, color = "navyblue") +
    geom_vline(xintercept = stat, linewidth = 1, linetype = 19,
             color = "purple") +
  geom_vline(xintercept = -stat, linewidth = 1, linetype = 19,
             color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = -stat, y = 2.25, color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = stat, y = 2.25, color = "purple") +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2.5)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal()
```
]
.panel[.panel-name[Calculations]
```{r}
xbar <- mean(race17$time_mins)
s <- sd(race17$time_mins)
n <- nrow(race17)
SE <- s/sqrt(n)
(t_stat <- (xbar - 93.29)/SE)
(crit_value99 <- qt(.995, df = n - 1))
(p_value <- 2*pt(t_stat, df = n - 1))
```
]]

---

## Hypothesis test for one mean

$$H_0: \mu_{2017} = 93.29$$
$$H_A: \mu_{2017} \neq 93.29$$

```{r}
t_test(race17, response = time_mins, 
       mu = 93.29, conf_level = 0.99)
```

We FAIL TO REJECT H0. That is, there is NOT sufficient evidence to claim that the average time is different in 2017, compared to 2006.

---

## What if we had more data?

We actually have a **census**

```{r}
library(cherryblossom)
glimpse(run17)
```

---

## What if we had more data?

```{r}
mean(run17$net_sec, na.rm = TRUE)
mean(run17$net_sec, na.rm = TRUE)/60
```

Runners ***are*** in fact faster in 2017, on average, compared to 2006, but we did not have enough evidence to support that conclusion based on 100 people. 

---

$$T = \frac{\overline{x} - \mu}{s/\sqrt{n}} = \frac{`r round(xbar, 2)` - 93.29}{`r round(s, 2)`/\sqrt{n}}$$

```{r, echo = FALSE}
xbar <- mean(race17$time_mins)
s <- sd(race17$time_mins)

n <- 100
stat <- abs((xbar - 93.29)/(s/sqrt(100)))

p1 <- ggplot() +
  xlim(c(-5,5)) + 
  stat_slab(aes(y = "T",
                   dist = dist_student_t(df = n - 1),
                   fill = after_stat(x < -stat | x > stat)),
               show.legend = FALSE)  +
  geom_vline(xintercept = qt(.995, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  geom_vline(xintercept = qt(.005, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.995, n - 1), y = 2, color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.005, n - 1), y = 2, color = "navyblue") +
    geom_vline(xintercept = stat, linewidth = 1, linetype = 19,
             color = "purple") +
  geom_vline(xintercept = -stat, linewidth = 1, linetype = 19,
             color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = -stat, y = 2.25, color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = stat, y = 2.25, color = "purple") +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2.5)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal() +
  ggtitle("n = 100")

n <- 1000
stat <- abs((xbar - 93.29)/(s/sqrt(n)))

p2 <- ggplot() +
  xlim(c(-5,5)) + 
  stat_slab(aes(y = "T",
                   dist = dist_student_t(df = n - 1),
                   fill = after_stat(x < -stat | x > stat)),
               show.legend = FALSE)  +
  geom_vline(xintercept = qt(.995, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  geom_vline(xintercept = qt(.005, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.995, n - 1), y = 2, color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.005, n - 1), y = 2, color = "navyblue") +
    geom_vline(xintercept = stat, linewidth = 1, linetype = 19,
             color = "purple") +
  geom_vline(xintercept = -stat, linewidth = 1, linetype = 19,
             color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = -stat, y = 2.25, color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = stat, y = 2.25, color = "purple") +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2.5)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal() +
  ggtitle("n = 1000")

n <- 5000
stat <- abs((xbar - 93.29)/(s/sqrt(n)))

p3 <- ggplot() +
  xlim(c(-5,5)) + 
  stat_slab(aes(y = "T",
                   dist = dist_student_t(df = n - 1),
                   fill = after_stat(x < -stat | x > stat)),
               show.legend = FALSE)  +
  geom_vline(xintercept = qt(.995, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  geom_vline(xintercept = qt(.005, n - 1), linewidth = 1, linetype = 21,
             color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.995, n - 1), y = 2, color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = qt(.005, n - 1), y = 2, color = "navyblue") +
    geom_vline(xintercept = stat, linewidth = 1, linetype = 19,
             color = "purple") +
  geom_vline(xintercept = -stat, linewidth = 1, linetype = 19,
             color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = -stat, y = 2.25, color = "purple") +
  annotate("label", label = "Test\nStatistic", 
           x = stat, y = 2.25, color = "purple") +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2.5)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal() +
  ggtitle("n = 5000")


grid.arrange(p1, p2, p3, ncol = 1)
```

---

## Sampling distribution of xbar

$\overline{x} = `r round(xbar, 2)`$ is not discernably different from $\mu = 93.29$ unless we have a large enough sample size. (Black curve: n = 5000; Red curve: n = 1000; Blue curve: n = 100)

```{r, echo = FALSE}
mu <- 93.29
ggplot() +
  stat_function(data = data.frame(x = c(80, 105)), aes(x),
                fun = dnorm, n = 101, 
                args = list(mean = mu, sd = s/sqrt(100)),
                color = "blue") + 
  stat_function(data = data.frame(x = c(80, 105)), aes(x),
                fun = dnorm, n = 101, 
                args = list(mean = mu, sd = s/sqrt(1000)),
                color = "red") + 
  stat_function(data = data.frame(x = c(80, 105)), aes(x),
                fun = dnorm, n = 101, 
                args = list(mean = mu, sd = s/sqrt(5000)),
                color = "black") + 
  geom_vline(xintercept = xbar, linewidth = 1, linetype = 20,
             color = "purple") +
  labs(y = "",
       x = "xbar",
       title = "Sampling distribution of xbar",
       subtitle = "when mu = 93.29") +
  scale_y_continuous(breaks = NULL) +
  theme_classic()
```

---

## So how big does n need to be, really?

Recall, we will reject at $\alpha = 0.01$ when $T = \frac{\overline{x} - \mu}{s/\sqrt{n}} \approx `r round(qnorm(.995), 3)`$

```{r}
qnorm(.995)
```

Solve for $n$!

$$n = \left(\frac{T*s}{\overline{x} - \mu}\right)^2$$

```{r}
n <- (2.576*s/(xbar - mu))^2
n
```

