---
title: "Exam 01"
subtitle: "Data Wrangling & Visualization"
date: "due Tuesday, October 1, 11:59pm"
output: 
  tufte::tufte_html:
    css: "hw.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

# Exam ground rules

-   This is an INDIVIDUAL EXAM. You may NOT work together or consult other people on this exam.
-   You are ENCOURAGED to use your notes, lecture slides/videos, previous assignments, and the internet, with some restrictions:
    -   You may utilize already existing resources on the internet, but you may NOT post to any online forums or ask an AI Chat bot (e.g. ChatGPT) to ask for help on the exam.\
    -   If you find code online, cite it, and annotate the code to demonstrate understanding.
-   Failure to adhere to any of the above policies will be considered a violation of academic integrity and will be handled according to university policy.

# Goals

This Exam will assess the following skills:

-   Data wrangling
-   Data visualization
-   Clarity of code and written communication
-   Ability to use data to investigate questions

# Getting started

-   Download `Exam1.qmd` from GitHub
-   Open `Exam1.qmd`

# Packages

We will use the following packages for this exam:

```{r load-packages, message = FALSE}
library(tidyverse)
library(dsbox)
library(ggridges)
library(kableExtra)
```

# Data

You will analyze two different datasets in this exam:

-   `avengers` (Exercises 1 - 5): data on 173 Marvel characters
-   `edibnb` (Exercises 6 - 8): data on 13,245 AirBnB reservations in Edinburgh, Scotland

# Avengers Data

This data was originally collected for [a FiveThirtyEight article](https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/).
The version of the `avengers` data we will work with here can be found in the `avengers.csv` file in the `data` folder you downloaded from our GitHub repo.
The code below will load the data.

```{r, eval = FALSE}
avengers <- read_csv("data/avengers.csv")
```

This dataset includes information about characters across the entire Marvel Cinematic Universe (MCU), so some of the names will be familiar if you are a fan of the films or comics.
Don't worry if you aren't a Marvel fan; no background knowledge is needed to successfully complete this exam!

We will focus on the following variables in this exam:

| Header                | Definition                                                                                                                                                                        |
|---------------------|---------------------------------------------------|
| `name`                | The full name or alias of the character                                                                                                                                           |
| `appearances`         | The number of comic books that character appeared in as of April 30                                                                                                               |
| `current`             | Is the member currently active on an avengers affiliated team?                                                                                                                    |
| `gender`              | The recorded gender of the character                                                                                                                                              |
| `probationary_introl` | Sometimes the character was given probationary status as an Avenger, this is the date that happened. The value will be *NA* if the character was never given probationary status. |
| `full_reserve`        | The month and year the character was introduced as a full or reserve member of the Avengers                                                                                       |
| `year`                | The year the character was introduced as a full or reserve member of the Avengers                                                                                                 |
| `years_since_joining` | 2015 minus the year                                                                                                                                                               |
| `death1`              | Yes if the Avenger died, No if not.                                                                                                                                               |
| `return1`             | Yes if the Avenger returned from their first death, No if they did not, blank if not applicable                                                                                   |

See FiveThirtyEight's [GitHub repo](https://github.com/fivethirtyeight/data/tree/master/avengers) for the full codebook.

1.  You are interested in creating a data frame with the most "classic" Avengers.
    Create a new data frame that only includes Avengers that were 1) created in 1970 or earlier and 2) were not given probationary status.
    Call the new data frame `classic_avengers`.
    *Hint: `is.na()` will be useful. Check: your new data frame should have 27 observations.*

2.  *Who are the newest classic Avengers?* Using a single pipeline: Create a new variable called `years_served` that represents the number of years served **as of 2024.** *(Hint: you can use either the `year` variable or `years_since_joining` variables to do this.)* Arrange the data appropriately to display the newest classic Avengers first.
    Display only necessary columns and the first 10 rows in a nicely formatted table.

    -   In addition to the table, report the names of the three newest classic Avengers and how long they have served in your narrative text.

3.  *Has the percentage of female Avengers changed over time?* To explore this question, compute the percentage of female Avengers in the `classic_avengers` dataset and compare it to the percentage of females among **all** Avengers.
    What do you conclude based on these results?

4.  Sort the full `avengers` dataset in descending order of `appearances` and display only the columns `name`, `appearances`, `death1`, and `return1` for the top five observations.

    -   What do you observe about these Avengers in terms of deaths and returns?

5.  *Do characters who die at least once tend to appear more or less often than characters who never die?* Create a visualization using the full `avengers` dataset to examine the distribution of appearances by whether or not character has died at least once.
    There are multiple correct way to visualize this, just make sure your visualization is clean and easy to read.

    -   What do you learn about Marvel movies from your results?

# Airbnb Data

The Airbnb data can be found in the **dsbox** package, and it's called `edibnb`.
Since the dataset is distributed with the package, we can load it with the following code:

```{r, eval = FALSE}
edibnb <- dsbox::edibnb
```

You can find out more about the dataset by inspecting its documentation, which you can access by running `?edibnb` in the Console or using the Help menu in RStudio to search for `edibnb`.
You can also find this information [here](https://rstudio-education.github.io/dsbox/reference/edibnb.html).

6.  Create a faceted histogram where each facet represents a neighborhood and displays the distribution of Airbnb prices in that neighbourhood. Think critically about whether it makes more sense to stack the facets on top of each other in a column, lay them out in a row, or wrap them around. Along with your visualisation, include your reasoning for the layout you chose for your facets.
7.  Use a single pipeline to identity the neighbourhoods with the top five (most expensive) median listing prices. Then, in another pipeline filter the data for these five neighbourhoods, and make ridge plots of the distributions of listing prices in these five neighbourhoods. In a third pipeline, calculate the minimum, mean, median, max, IQR, and standard deviation of listing price in each of these neighbourhoods. Use the visualization and the summary statistics to describe the distribution of listing prices in the neighbourhoods. (Your answer will include three pipelines - one of which ends in a visualization - and a narrative.)
8.  Create a visualization that will help you compare the distribution of review scores (`review_scores_rating`) across neighbourhoods. You get to decide what type of visualisation to create and there is more than one correct answer! In your answer, include a brief interpretation of how Airbnb guests rate properties in general and how the neighbourhoods compare to each other in terms of their ratings.

# Further exploration

9.  Choose one of the two datasets - `avengers` or `edibnb` - and comment on each of the following:

-   What you found through this analysis

-   What limitations might exist to the data/analyses

-   What additional data you would like to have to explore further questions

# BONUS

State an additional research question you could investigate using either the `avengers` or `edibnb` data.
State the question, provide a visualization that investigates it, and comment on what your visualization shows.

# Submission

Before submitting, open up your .html in a full window and inspect the following:

-   Check your code for neatness - add spaces and line breaks where appropriate to improve readability

-   Check any visualizations for informative and clean labels and titles

-   Check any tables for clean formatting (e.g. with `kbl() |> kable_minmal()`)

-   Check that you've interpreted output as appropriate

-   Make sure extraneous warnings and messages are suppressed (e.g. set `#| warning: false`, `#| message: false`)

To submit, upload your .html file to the Exam 1 assignment on Blackboard.
Refer back to [Lab 01](https://kgfitzgerald.github.io/APU_MATH_250/labs/lab-01-hello-r.html) for detailed instructions on how to export and submit your assignment.

# Grading (50 pts)

<br>

| Component          | Points |
|:-------------------|:-------|
| Ex 1               | 5      |
| Ex 2               | 5      |
| Ex 3               | 5      |
| Ex 4               | 5      |
| Ex 5               | 5      |
| Ex 6               | 5      |
| Ex 7               | 5      |
| Ex 8               | 5      |
| Ex 9               | 5      |
| Reflection prompts | 5      |
| BONUS              | 2      |
