---
title: "Hypothesis testing"
subtitle: "STAT 4380"
author: "Katie Fitzgerald, adpated from datasciencebox.org"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(DT)
library(emo)
library(openintro)
library(ggdist)
library(distributional)
conflicts_prefer(infer::generate)
```

class: middle

# Hypothesis testing for a regression slope

## Is there a relationship between x and y?

### Is $\beta_1 \neq 0$?

$$\hat{y} = \beta_0 + \beta_1x$$

---

```{r message=FALSE, echo = FALSE}
library(tidyverse)
library(openintro)
```

## Data & model

```{r message=FALSE}
m1 <- lm(gift_aid ~ family_income, data = elmhurst)
tidy(m1)
```

Once we've taken uncertainty into account, is $b_1$ different enough from zero to claim that there is a true relationship between family income and gift aid?

---

## Parameter vs. statistic

A **parameter** for a hypothesis test is the "true" value of interest. We typically estimate the parameter using a **sample statistic** as a **point estimate**.

$\beta_1~$: true relationship between gift aid and family income

$b_1~$: relationship (slope) between gift aid and family income in the sample = `r round(tidy(m1)[2,2],2)`



---

## Two claims

- **Null hypothesis:** "There is nothing going on"

There is no relationship between gift aid and family income $(\beta_1 = 0)$

--

- **Alternative hypothesis:** "There is something going on"

There is a relationship between gift aid and family income $(\beta_1 \neq 0)$

---

## Hypothesis testing as a proof by contradiction

In order to prove A, 
+ Start by assuming A is false
+ Analyze the logical consequences of A being false, then
+ If you run into a contradiction, that is evidence that your premise (A = false) was incorrect. That is, you now have evidence that A = true

In the language of hypothesis testing, we want to prove $H_A$. So, 
+ Start by assuming $H_0$ is true (i.e. that $H_A$ is false)
+ Analyze the logical consequences of $H_0$ being true 
    + What would the sampling distribution of the test statistic look like when $H_0$ is true?
+ If your observed test statistic is unlikely to have happened if $H_0$ is true, then reject $H_0$ in favor of $H_A$

---

## Hypothesis testing as a court trial

- **Null hypothesis**, $H_0$: Defendant is innocent

- **Alternative hypothesis**, $H_A$: Defendant is guilty

--

- **Present the evidence:** Collect data

--

- **Judge the evidence:** "Could these data plausibly have happened by chance if the null hypothesis were true?"
    * Yes: Fail to reject $H_0$
    * No: Reject $H_0$
    
---

## Sampling distributions & test statistics

$$H_0: \beta_1 = 0$$
$$H_A: \beta_1 \neq 0$$

Mathematical theory tells us the when $H_0$ is true, 
$$T = \frac{b_1 - \beta_1}{SE(b_1)} = \frac{b_1}{SE(b_1)} \sim t(df = n - 2)$$
```{r, echo = FALSE}
tidy(m1)
```

```{r}
-0.0431/0.0108
```

---

## Compare test statistic to reference distribution

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "t-plot", echo = FALSE, warning = FALSE, out.height = "60%"}
```
]
.panel[.panel-name[Code]
```{r t-plot, fig.show="hide"}
t_stat <- abs(tidy(m1)[2,2]/tidy(m1)[2,3]) |> pull()
crit_value <- qt(.975, nrow(elmhurst) - 2)
ggplot() +
  xlim(c(-5,5)) + 
  stat_halfeye(aes(y = "t(48)", color = "pink",
                   dist = dist_student_t(df = 48),
                   fill = after_stat(x < -4 & x > 4)),
               show.legend = FALSE) +
  geom_vline(xintercept = t_stat, linewidth = 1, linetype = 20) +
  geom_vline(xintercept = -t_stat, linewidth = 1, linetype = 20) +
  annotate("label", label = "Test\nStatistic", x = t_stat, y = 1.5) +
  annotate("label", label = "Test\nStatistic", x = -t_stat, y = 1.5) +
  geom_vline(xintercept = crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  geom_vline(xintercept = -crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  annotate("label", label = "Critical\nValue", x = crit_value, y = 1.75) +
  annotate("label", label = "Critical\nValue", x = -crit_value, y = 1.75) +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal()
```
]
]


---

## Hypothesis testing framework

- Start with a null hypothesis, $H_0$, that represents the status quo

- Set an alternative hypothesis, $H_A$, that represents the research question, i.e. what we’re testing for

- Conduct a hypothesis test under the assumption that the null hypothesis is true
    + If test statistic is MORE extreme than the critical value, REJECT the null hypothesis in favor of the alternative (data provide convincing evidence the null hypothesis is false)
    + If test statistic is LESS extreme than the critical value, FAIL TO REJECT the null hypothesis (data do NOT provide convincing evidence the null hypothesis is false)

- **p-value** is the probability of observing a result as extreme or more extreme than what was observed, given that the null hypothesis is true

---

## Hypothesis testing framework

+ In other words, the **p-value** is the probability that falls BEYOND the test-statistic
+ REJECT $H_0$ for small p-values
+ The probability that falls beyond the critical value will always be $\alpha$
+ Reject when p-value $< \alpha$


```{r, echo = FALSE, fig.height="40%"}
ggplot() +
  xlim(c(-5,5)) + 
  stat_halfeye(aes(y = "t(48)", color = "pink",
                   dist = dist_student_t(df = 48),
                   fill = after_stat(x < -4 & x > 4)),
               show.legend = FALSE) +
  geom_vline(xintercept = t_stat, linewidth = 1, linetype = 20) +
  geom_vline(xintercept = -t_stat, linewidth = 1, linetype = 20) +
  annotate("label", label = "Test\nStatistic", x = t_stat, y = 1.5) +
  annotate("label", label = "Test\nStatistic", x = -t_stat, y = 1.5) +
  geom_vline(xintercept = crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  geom_vline(xintercept = -crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  annotate("label", label = "Critical\nValue", x = crit_value, y = 1.75) +
  annotate("label", label = "Critical\nValue", x = -crit_value, y = 1.75) +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal()
```

---

## Hypothesis Test Conclusion

$$\widehat{\text{gift_aid}} = \beta_0 + \beta_1\text{family_income}$$

$$H_0: \beta_1 = 0$$
$$H_A: \beta_1 \neq 0$$

```{r}
tidy(m1)
```

Because our `statistic` (`-3.98`) is more extreme than the critical value (or equivalently, because our p-value  = `2.29e-4` < 0.05), we REJECT $H_0$. That is, there is sufficient evidence to claim there is a statistically significant relationship between family income and gift aid. 


---

## Significance level $\alpha$

Often, 5% is used as the cutoff for whether the p-value is low enough that the data are unlikely to have come from the null model. This cutoff value is called the **significance level**, $\alpha$.

- If p-value < $\alpha$, reject $H_0$ in favor of $H_A$: The data provide convincing evidence for the alternative hypothesis.

- If p-value > $\alpha$, fail to reject $H_0$ in favor of $H_A$: The data do not provide convincing evidence for the alternative hypothesis.

$\alpha$ controls the Type I error rate - the probability of falsely rejecting the null hypothesis when it's true (false positive). Often, smaller $\alpha$ may be appropriate (e.g. 0.01, 0.001).

Decreasing $\alpha$ inherently increases Type II error (false negative)

---

## Significance level $\alpha$

```{r, echo = FALSE, fig.height="40%"}
crit_value99 <- qt(.995, nrow(elmhurst) - 2)
ggplot() +
  xlim(c(-5,5)) + 
  stat_halfeye(aes(y = "t(48)", color = "pink",
                   dist = dist_student_t(df = 48),
                   fill = after_stat(x < -4 & x > 4)),
               show.legend = FALSE) +
  geom_vline(xintercept = crit_value99, linewidth = 1, linetype = 21,
             color = "navyblue") +
  geom_vline(xintercept = -crit_value99, linewidth = 1, linetype = 21,
             color = "navyblue") +
    geom_vline(xintercept = crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  geom_vline(xintercept = -crit_value, linewidth = 1, linetype = 21,
             color = "pink") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = crit_value99, y = 2, color = "navyblue") +
  annotate("label", label = "0.01 Critical\nValue", 
           x = -crit_value99, y = 2, color = "navyblue") +
  annotate("label", label = "0.05 Critical\nValue", 
           x = crit_value, y = 1.75, color = "pink") +
  annotate("label", label = "0.05 Critical\nValue", 
           x = -crit_value, y = 1.75, color = "pink") +
  coord_cartesian(expand = FALSE, ylim = c(0.9, 2.5)) +
  labs(y = "", x = "x") +
  scale_fill_viridis_d() +
  theme_minimal()
```

---

## Setting the hypotheses

.question[
Which of the following is the correct set of hypotheses?
]

(a) $H_0: \beta_1 = 0$; $H_A: \beta_1 \neq 0$ <br>

(b) $H_0: \beta_1 = 0$; $H_A: \beta_1 < 0$ <br>

(c) $H_0: \beta_1 = 0$; $H_A: \beta_1 > 0$ <br>

(d) $H_0: b_1 = 0$; $H_A: b_1 \neq 0$ <br>

(e) $H_0: b_1 = 0$; $H_A: b_1 < 0$ <br>

(f) $H_0: b_1 = 0$; $H_A: b_1 > 0$ <br>

---

class: middle

# One vs. two sided hypothesis tests

---

## Types of alternative hypotheses

- One sided (one tailed) alternatives: The parameter is hypothesized to be less than or greater than the null value, < or >
    - p-value calculated as probability that falls beyond the test statistic in ONE TAIL

--

- Two sided (two tailed) alternatives: The parameter is hypothesized to be not equal to the null value, $\ne$
    - p-value calculated as two times the tail area beyond the observed test statistic (i.e. probability more extreme in BOTH TAILS)
    - More objective, and hence more widely preferred
    
---

## Confidence Interval vs. Hypothesis test

+ 95% confidence interval $\iff$  $\alpha = 0.05$ hypothesis test
+ 99% confidence interval $\iff$  $\alpha = 0.01$ hypothesis test 
+ 90% confidence interval $\iff$  $\alpha = 0.10$ hypothesis test 

$$H_0: \beta_1 = 0$$
$$H_A: \beta_1 \neq 0$$

Rejecting $H_0$ $\iff$ CI does NOT include 0

Failing to reject $H_0$ $\iff$ CI DOES include 0

---

## Confidence Interval vs. Hypothesis Test

What do you want to do?

- Estimation -> Confidence interval
    + Magnitude of parameter is of interest
- Decision -> Hypothesis test
    + Is there enough evidence to conclude ____? 
    + E.e. drug works, curriculum is effective, defendant is guilty
