---
title: "Week 05"
subtitle: "Scientific Practice & Data Storytelling"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(tidyverse)
```


### Statistician(s) of the Week

[Maricela Cruz](https://hardin47.github.io/CURV/scholars/cruz.html)

[Alejandra Castillo](https://hardin47.github.io/CURV/scholars/castillo.html)

### Announcements

+ Sign up for project group in Blackboard if you haven't already - sit with your partner today
+ In-class exam made on Thursday, will also release take-home exam due next Tuesday at midnight
    + AE solutions have been posted (home page of Blackboard)
    + everything graded - reminder if you submitted on time, can revise & resubmit for up to full credit w/in one week of when assignment was graded
+ No lab or annotations this week, but you will have lecture recordings & prep quiz for Tuesday (~60 mins, 3 videos)
+ Project proposal due next Saturday


<!-- ### Exam ground rules -->

<!-- + This is an INDIVIDUAL EXAM. You may NOT work together or consult other people on this exam. -->
<!-- + You are ENCOURAGED to use your notes, lecture slides/videos, previous assignments, and the internet -->
<!--     + You may utilize already existing resources on the internet, but you may NOT post to any online forums or ask as AI chat bot for help on the exam.   -->
<!--     + If you find code online, cite it, and annotate the code to demonstrate understanding.  -->
<!-- + Failure to adhere to any of the above policies will be considered a violation of academic integrity and will be handled according to university policy. -->

<!-- ### Reminders / Tips -->

<!-- + Make sure you've answered all parts of the question! -->
<!-- + Make sure to clean up visualizations - add titles, clean axis labels, all text should be readable -->
<!-- + Use `kbl()` for clean formatting of tables -->
<!-- + Google your errors -->
<!-- + Search "how to ____ in tidyverse r" -->
<!-- + stackoverflow forum is usually particularly helpful -->
<!-- + Check out textbooks linked in prep material (esp. R for Data Science) -->

### Questions?

<!-- + How does the `if_else()` statement work? -->
<!-- + When should I save the output of a pipe back into a dataset? -->

<!-- ### Project Work Session -->

<!-- - See [Doing Data Science slides](https://kgfitzgerald.github.io/NOVA_STAT_4380/slides/16-doing-data-science.html#1) for getting started on your [project proposal](https://apumath250.netlify.app/project/#project-proposal) -->


