---
title: "Lab 08 - Working with Strings"
subtitle: "Babynames"
format:
  html:
    css: styles.css
    grid:
      margin-width: 350px
fig-numbering: false
toc: true
toc-location: right
categories:
  - strings
  - counts & proportions
  - data wrangling
  - data visualization
editor_options:
  chunk_output_type: console
callout-icon: false
---

```{r include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction

In this lab, you will use string functions to investigate meaningful questions about naming patterns in the United States over time. 

## Learning goals

By the end of this lab, you should be able to:

+ Use `stringr` verbs such as `str_detect()`, `str_starts()`, `str_ends()`, `str_sub()`, `str_length()`
+ Create new variables based on string patterns
+ Distinguish between different units of observation
+ Decide when to use counts versus proportions
+ Create visualization to communicate trends over time

## Load packages and data

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(babynames)

data(babynames)
```

We will be using the `babynames` data from the `babynames` package, which includes counts of babynames in the United States from 1880 to 2017. A name is included if it occurs at least 5 times in that calendar year. The data come from the U.S. Social Security Administration. 

**Note: each row contains COUNTS and PROPORTIONS for a name\*sex\*year combination. Each row does NOT represent one baby. Keep this in mind as you analyze the data.**

# Exercises

::: callout-note
### Exercise 1

Create four new variables in the `babynames` data:

+ `name_length` (counts the number of letters in the name)
+ `first_letter` 
+ `last_letter`
+ `name_ending` (extract the last three letters)

:::

::: callout-note
### Exercise 2

+ How many unique names contain the string "liz"?
+ Produce a table with counts of the top 10 variations of "liz". *Hint: you should sum over all years first*

:::

::: callout-note

### Exercise 3

*Are girl names more likely to end in vowels (aeiouy)?*

+ Create a variable indicating whether the name ends in a vowel
+ For each year and sex, what proportion of babies received a vowel-ending name? Has that changed over time? Is the pattern different for boys and girls?
    + Create a line plot to investigate
    + Briefly comment on your results
:::

::: callout-note
### Exercise 4
*Have names starting with K become more common?*

+ Produce a visualization that investigates this by sex. Comment on your results. 
+ Choose two additional letters to investigate, and provide a 2nd plot that shows the trends over time for the three letters. 

:::

::: callout-note
### Exercise 5
*Are longer names becoming more common?*

+ Plot the average name length by sex over time

:::

::: callout-note
### Exercise 6

*What is the most common letter of first names? Has this changed over time? Does this differ by sex?*

Produce an appropriate visualization to explore this. Briefly comment on your results. *Hint: it may be helpful to brainstorm a useful visualization by sketching by hand first*

:::

::: callout-note
### Exercise 7

*Which name endings (last three letters) are most popular among boys versus girls?* 

Produce a table that shows the top 5 name endings for each sex. 

:::

::: callout-note
### Exercise 8

Create a visualization that explores the popularity of your name over time. Briefly comment on the results.

:::

::: callout-note
### BONUS

Propose one additional question that can be investigated with these data, and provide a visualization that investigates it.

:::








