---
title: "Lab 06 - Grading the professor, Pt. 2"
subtitle: "Modelling with multiple predictors"
date: "due Thursday, October 24 at 4:20pm"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

```{r include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

In this lab we revisit the professor evaluations data we modelled in the previous AE.
Last time, we modeled evaluation scores using a single predictor at a time.
This time we will use multiple predictors to model evaluation scores.

For context, review the previous AE's introduction before continuing on to the exercises.

# Learning goals

-   Fitting a linear regression with multiple predictors
-   Interpreting regression output in context of the data
-   Comparing models

# Getting started

+ You will be working in your randomly assigned Lab 07 - 08 Groups. See Blackboard.
+ Open our class [GitHub repo](https://github.com/kgfitzgerald/NOVA_STAT_4380/tree/main/assignment_templates) and download the .qmd file for this lab. Refer back to [Lab 01](https://kgfitzgerald.github.io/NOVA_STAT_4380/labs/lab-01-hello-r.html) for detailed instructions on how to get started on a lab. 

All plots should follow the best visualization practices discussed in lecture, including an informative title, labeled axes, and careful consideration ofaesthetic choices.

All code should follow the tidyverse style guidelines.

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualisation, and the data lives in the **openintro** package.

```{r}
library(tidyverse) 
library(openintro)
library(broom)
```

## Data

The data can be found in the **openintro** package, and it's called `evals`.
Since the dataset is distributed with the package, we don't need to load it separately; it becomes available to us when we load the package.
You can find out more about the dataset by inspecting its documentation, which you can access by running `?evals` in the Console or using the Help menu in RStudio to search for `evals`.
You can also find this information [here](https://www.openintro.org/data/index.php?data=evals).

# Exercises

## Simple linear regression

1.  Fit a linear model (one you have fit before): `score_bty_fit`, predicting average professor evaluation `score` based on average beauty rating (`bty_avg`) only. Write the linear model, and note the $R^2$ and the adjusted $R^2$.

## Multiple linear regression

2.  Fit a new linear model: `score_bty_gen_fit`, predicting average professor evaluation `score` based on average beauty rating (`bty_avg`) and `gender`.
    Write the linear model, and note the $R^2$ and adjusted $R^2$.

The following code plots the above regression model. Include this plot in your report. 

```{r, echo = FALSE}
score_bty_gen_fit <- linear_reg() |> 
  set_engine("lm") |> 
  fit(score ~ bty_avg + gender, data = evals)
```


```{r}
#create augmented dataset
score_bty_gen_fit_aug <- augment(score_bty_gen_fit)

#plot data + regression line
ggplot(score_bty_gen_fit_aug, aes(y = score, x = bty_avg, 
                                  color = gender)) +
  geom_point() +
  geom_line(aes(y = .fitted), size = 1.5)
```


3.  Interpret the slopes and intercept of `score_bty_gen_fit` in context of the data.

4.  What percent of the variability in `score` is explained by the model `score_bty_gen_fit`? *Use inline code to include your answer in a sentence. Convert to a percentage and then round to 1 decimal place.*

5.  What is the equation of the line corresponding to *just* male professors? What is it for female professors? 

6.  For two professors who received the same beauty rating, which gender tends to have the higher course evaluation score?

7.  How do the adjusted $R^2$ values of `score_bty_gen_fit` and `score_bty_fit` compare? What does this tell us about how useful `gender` is in explaining the variability in evaluation scores when we already have information on the beauty score of the professor. *Use inline code to include the numbers in your sentence(s).*

8. Compare the slopes of `bty_avg` under the two models (`score_bty_fit` and `score_bty_gen_fit`). Has the addition of `gender` to the model changed the parameter estimate (slope) for `bty_avg`? Discuss what this might indicate. *Use inline code to include the numbers in your sentence(s).*

9. Create a new model called `score_bty_rank_fit` with `gender` removed and `rank` added in, and provide a visualization of the data with the relevant regression line(s). Write the equation of the linear model, interpret the slopes and intercept in context of the data. *Use inline code to include the numbers in your sentence(s).*
    
## Interaction models

How does the relationship between beauty and evaluation score vary between male and female professors? 

10. To answer the above question, do the following:

    + Fit an interaction model (with an interaction between `bty_avg` and `gender`)
    + Provide a visualization of the data with the relevant regression lines. 
    + Write the equation of the linear model separately for males and females
    + Interpret the slopes and intercepts in the context of the data. 
    + Decide which model is more appropriate in this case - the main effects or interaction model. Justify your choice. 


# Submission

Before submitting your assignment on Blackboard, Render one last time, then open up your .html in a full window and inspect the following:

+ Check your code for neatness - add spaces and line breaks where appropriate to improve readability

+ Check any visualizations for informative and clean labels and titles

+ Check any tables for clean formatting (e.g. with `kbl() |> kable_minmal()`)

+ Check that you've interpreted output as appropriate

+ Make sure extraneous warnings and messages are suppressed (e.g. set `#| warning: false`, `#| message: false`)

+ Check that Exercises are clearly labeled and your text responses are visually distinguished

+ Check for overall neatness & organization of the report (e.g. headers, subheaders, use of bullets, font changes)

To submit, Render one last time and then upload your .html document to the relevant assignment on Blackboard. Refer back to [Lab 01](https://kgfitzgerald.github.io/NOVA_STAT_4380/labs/lab-01-hello-r.html) for detailed instructions on how to submit your assignment. 

# Grading (50pts)

| Component | Points |
|:----------|:-------|
| Ex 1      | 2      |
| Ex 2      | 2    |
| Ex 3      | 4   |
| Ex 4      | 2   |
| Ex 5      | 4      |
| Ex 6      | 2    |
| Ex 7      | 4   |
| Ex 8      | 4   |
| Ex 9      | 8   |
| Ex 10  | 8 |
| Reflection prompts    | 5     |
| Workflow & formatting | 5     |

**Grading notes**:

-   The "Workflow & formatting" grade is to assess the reproducible workflow. This includes having readable code (e.g. adequate use of spacing and line breaks), labeled code chunks, informative headers and sub-headers, and an overall organized and uncluttered report (e.g. suppress messages & warnings, no extraneous output).
