---
title: "Lab 02 – Data Visualization"
format:
  html:
    css: styles.css
    grid:
      margin-width: 350px
    self-contained: true
fig-numbering: false
toc: true
toc-location: right
categories:
  - Data visualization
  - Tidyverse
  - Ridge plots
  - ANES
editor_options:
  chunk_output_type: console
callout-icon: false
---

# Goals

In this lab, you will:

+ Gain proficiency in data visualization
+ Apply principles of effective visualization to a real dataset
+ Continue developing a workflow for reproducible data analysis

# Getting Started

+ Go to our class [GitHub repo](https://github.com/kgfitzgerald/NOVA_STAT_4380/tree/main/assignment_templates) and download the `.qmd` file for this lab.
+ Refer back to [Lab 01](https://kgfitzgerald.github.io/NOVA_STAT_4380/labs/lab-01-hello-r.html) for instructions on how to get started on a lab.
+ You will work in your Lab 01–02 groups (see Brightspace).

# Packages

We will work with the `tidyverse` package as usual. We will also use `viridis` and the `ggridges` packages.

```{r load-packages, message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse)
library(viridis)
library(ggridges)
```

# Data: 2020 American National Election Study

The dataset comes from the
[2020 American National Election Study](https://electionstudies.org/data-center/2020-time-series-study/).

```{r, eval = FALSE}
anes <- read_csv("data/anes2020_subset.csv")
```

A subset of variables are provided here. Some of them have been recoded, while others you may need to recode in order to be able to carry out your analysis. The variables are as follows:

- `CASEID`: a Case ID for the respondent.

- `hunt_fish`: a dummy variable asking if the respondent has gone hunting or fishing in the past year.

- `scientists`: A feeling thermometer question that asks how warmly respondents feel towards scientists. A score of 0 represents the coolest rating, while a score of 100 represents the warmest rating.

- `education`: An ordinal categorical representing the highest level of education for the respondent, ranging from less than high school to a professional degree.

- `ideology`: a seven point self-rating scale for the respondent’s ideology ranging from most liberal to most conservative

- `urbanrural`: a variable indicating how rural or urban the respondent’s home community is with four possible values: rural, small town, suburb, or city.

# Exercises

All plots should follow the best visualization practices discussed in lecture. Plots should include an informative title, axes should be labeled, and careful consideration should be given to aesthetic choices.

Remember that continuing to develop a sound workflow for reproducible data analysis is important as you complete the lab and other assignments in this course. You should provide adequate comments to your code and narrative text interpreting your results. 

::: callout-note
### Exercise 1
Provide a brief description of the data collection process — who collected the data, when, who was surveyed, and anything else you find relevant/interesting.
:::

::: callout-note
### Exercise 2
How many rows and columns are in the `anes` dataset?
Include code and output.
:::

::: callout-note
### Exercise 3

Create a bar chart of `ideology` (counts on y-axis).

  - Include axis labels and an informative title.
  - What is the most common ideology? 
  - Do respondents tend to be moderate or more ideologically extreme?

:::

::: callout-note
### Exercise 4
Now, let’s examine whether ideologies are different based upon where people live. Please make a filled bar plot, showing one bar for each `ideology`, with the proportion of respondents on the y-axis going from 0-1, and the `fill` determined by `urbanrural`.

- Include labels
- Encouraged to use `viridis` colors

- Where do people of each ideology tend to live?
- Does the level of non-response vary by ideology?

:::

::: callout-note
### Exercise 5
*How do people view scientists?* 

+ Make a histogram of the `scientists` feeling thermometer. 
+ Comment on the distribution, including features such as skewness and peaks. Interpret in context.
:::

---

::: callout-note
### Exercise 6
*Does the ideology of those who have gone hunting or fishing in the past year differ from those who haven’t?* Explore this using **side-by-side boxplots**.

Start your code with:

```{r, eval = FALSE}
anes |>
  drop_na(hunt_fish) |>
  mutate(hunted_fished = ifelse(hunt_fish == 0,
                                "Did Not Hunt or Fish",
                                "Hunted or Fished"))
```

:::

::: callout-note
### Exercise 7
Explore the same question as in Exercise 6, this time using `geom_density_ridges()` to construct a ridge plot as an alternative to side-by-side box plots.
:::

::: {.marginfigure}
You can read more about ridge plots
[here](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html).
:::

::: callout-note
### Exercise 8
Describe what you observe in the boxplots (Ex 6) and ridge plots (Ex 7).
What can you learn from one plot that you do not see in the other or that adds additional context to the other?
:::

::: callout-note
### Exercise 9
Is education related to views of scientists? Create a scatterplot with:

- education on x-axis
- scientists feeling thermometer on y-axis
- add a best-fit line layer: `geom_smooth(method = "lm")`

Is this visualization useful? Why or why not?

:::

::: callout-note
### Exercise 10
What is an additional question you could investigate using this data? State the question, provide a visualization that investigates it, and comment on what your visualization shows.
:::

::: callout-note
### Exercise 11

Write a brief paragraph summarizing what you found in this lab, what limitations might exist to the data/analyses, and what additional data you would like to have to explore further questions.

:::


Before submitting your lab:

+ Render one last time, check the `.html` file for accuracy
+ Check your code for neatness - add spaces and line breaks where appropriate to improve readability
+ Check visualizations for clean titles and labels
+ Suppress extraneous messages/warnings (e.g. set `#| warning: false`, `#| message: false` inside code chunks)
+ Ensure exercises are clearly labeled and your text responses are visually distinguished 
+ Confirm neat organization and readable structure

# Submit

::: {.callout-important}

Unfortunately, Brightspace does not handle .html files well. Before submitting, you will need to convert your .html file to a .zip folder. Locate the file on your computer (not in RStudio), right click, and select "Compress." Then, upload the .zip folder to Brightspace.

:::



# Grading (10pts)

Labs are opportunities to practice the skills and habits you will use on larger course projects. They are graded holistically rather than point-by-point.

Your lab will be evaluated for:

* **Completion & substantive correctness (8 points):** You have attempted all exercises, your code and results are generally correct, and your written responses appropriately interpret and engage with the data.
* **Reproducible workflow & presentation (2 points):** Your document renders successfully and your code, output, visualizations, and responses are organized and readable.

Minor mistakes will not be penalized individually. You may also receive feedback about misconceptions or habits to improve even when they do not result in a point deduction. Use this feedback in future labs and projects.


### Lab rubric - 10 points total

#### Completion & substantive correctness — 8 points

| Score | Description |
|---|---|
| **8** | Complete and substantively correct. Code and analyses accomplish what was asked, and written responses appropriately interpret the results. Minor imperfections are okay. |
| **7** | Mostly complete and correct, with one notable error, omission, or misunderstanding. |
| **6** | Several errors or omissions, but the work demonstrates reasonable understanding of the main ideas. |
| **4–5** | Significant errors, omissions, or misunderstandings that should be addressed before moving on. |
| **0–3** | Substantially incomplete, incorrect, or shows limited engagement with the lab. |

#### Reproducible workflow & presentation — 2 points

| Score | Description |
|---|---|
| **2** | Document renders successfully; code, output, visualizations, and written responses are reasonably organized and readable. |
| **1** | Some problems with reproducibility, organization, readability, or presentation. |
| **0** | Major problems with reproducibility, organization, readability, or presentation. |

**Total: 10 points**

